---
title: 'Class 12: Homework'
author: 'Snehita Vallumchetla (PID: A16853399)'
date: "2025-02-15"
output: pdf_document
---

>Q13:Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

The sample size for the A/A genotype is 108 individuals with median expression levels of 31.2.
The sample size for the A/G genotype is 233 individuals with median expression levels of 25.1.
The sample size for the G/G genotype is 121 individuals with median expression levels of 20.1.

(my work is shown below)

```{r}
df <- read.table("expression.txt", header = TRUE, sep = "", stringsAsFactors = FALSE)
head(df)
```

```{r}
summary(df)
```
```{r}
table(df$geno)
```

```{r}
aa <- boxplot(exp ~ geno, data = df[df$geno == "A/A", ], main = 'expression of A/A')
ag <- boxplot(exp ~ geno, data = df[df$geno == "A/G", ], main = 'expression of A/G')
gg <- boxplot(exp ~ geno, data = df[df$geno == "G/G", ], main = 'expression of G/G')
```
```{r}
aa$stats
ag$stats
gg$stats
```

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

The median expression levels of the A/A genotype is greater than the median expression levels of the G/G phenotype. Since there is a difference in median expression levels between the A/A and G/G genotypes, this is indicates that the SNP does effect the expression of ORMDL3, likely lowering its expression. 

```{r}
boxplot(exp ~ geno, data = df,  
        main = "Expression by Genotype", 
        xlab = "Genotype", 
        ylab = "Expression",
        col = c("lightpink", "lightgreen", "lightblue"))
```






