---
title: "Comparative Structure Analysis"
author: "Snehita Vallumchetla (PID: A16853399)"
format: html
---

Here we run through a complete "pipeline" of structural analysis that begins with a single sequence indentifer and ends in a PCA analysis

```{r}
library(bio3d)
```

## Step 1

Retrieve a sequence for the protein we are interested in. We will take ADK "1ake_A"
```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```

## Step 2
Run a blast search of the PDB for all related sequences to find our input `aa`

```{r}
blast <- blast.pdb(aa)
```

## Step 3: Download all structures 
These are our "top hit" i.e. all the structures in the PDB database related to our input sequence. 

```{r}
hits <- plot(blast)
```


```{r}
hits$pdb.id
```


```{r}
files <- get.pdb(hits$pdb.id, path = 'pdbs', split = TRUE, gzip = TRUE)
```

## Step 4. Align and Superpose

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile = "msa")
```

```{r}
pdbs
```


## Step 5: PCA

Let's use our old friend PCA to make sense of these confusing, complicated, structure relationship. 

```{r}
pc <- pca(pdbs)
plot(pc)
```

Let's make a trajectory (or movie) of the main conformational changes captured by PC1. We will use the `mktrj()` function for this...

```{r}
mktrj(pc, file = 'pca_result.pdb')
```

Back of the envelope comparison of the PDB size to UniProt

```{r}
uniprot <- 253206171

pdb <- 231029

pdb/uniprot * 100
```


